Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-05-22 ms.tex
Use lowercase symbols for signals

A laser interferometer senses \acp{GW} by transducing differential changes in arm length to a digitized photocurrent time series. In the angular frequency domain, a single detector's observation $X_i(\omega)$ is
%
\begin{equation}\label{eq:signal-model}
	X_i (\omega) = H_i (\omega; \boldsymbol\theta) + N_i (\omega),
\end{equation}
%
where $H_i (\omega; \boldsymbol\theta)$ is the \ac{GW} signal receied by detector $i$, given a parameter vector $\boldsymbol\theta$ that describes the \ac{GW} source, and $N_i (\omega)$ is that detector's \ac{WSS} Gaussian noise with \ac{PSD} $S_i(\omega)$. We shall denote the combined observation from a network of detectors as $\mathbf X (\omega) \equiv \{X_i (\omega)\}_i$.

The likelihood, or the probability of obtaining the observation $\mathbf X$ conditioned on the value of $\boldsymbol\theta$, is Gaussian, with the log\nobreakdashes-likelihood proportional to:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\mathcal{L}(\mathbf X; \boldsymbol\theta) = \prod_i p(X_i | \boldsymbol\theta)
		\propto \exp \left[
		- \frac{1}{2} \sum_i \int_0^\infty \frac{\left|X_i (\omega)
			- H_i(\omega; \boldsymbol\theta) \right|^2}{S_i(\omega)} \, \mathrm{d}\omega
	\right].
\end{equation}

@@

Assuming circular orbits and no spin precession, we can write the \ac{GW} signal in each detector as a linear combination of two basis waveforms $H_0$ and $H_{\pi/2}$. For nonprecessing systems, $H_0$ and $H_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. In other words, $H_{\pi/2}(\omega) = i H_0(\omega)$ for all $\omega$. This linear combination isolates all dependence on the extrinsic parameters $\boldsymbol\theta_\mathrm{ex}$ into the coefficients and all dependence on the intrinsic parameters $\boldsymbol\theta_\mathrm{in}$ into the basis waveform,
%
\begin{eqnarray}\label{eq:full-signal-model}
	\fl H_i(\omega; \boldsymbol\theta) &=& e^{-i\omega t_\oplus}\frac{D_{1,i}}{D_\mathrm{L}}
	\left(\begin{array}{cc}
		F_{+,i}(h_i, \delta) \\
		F_{\times,i}(h_i, \delta)
	\end{array}\right)^\intercal
	\left(\begin{array}{cc}
		\cos{2\psi} & -\sin{2\psi} \\
		\sin{2\psi} & \cos{2\psi}
	\end{array}\right)
	\left(\begin{array}{cc}
		\frac{1}{2}\left(1+\cos^2{\iota}\right) & 0 \\
		0 & \cos{\iota}
	\end{array}\right) \nonumber\\\fl &\cdot&
	\left(\begin{array}{cc}
		\cos{2\phi_c} & -\sin{2\phi_c} \\
		\sin{2\phi_c} & \cos{2\phi_c}
	\end{array}\right)
	\left(\begin{array}{cc}
		1 \\
		i
	\end{array}\right) H_0(\omega; \boldsymbol\theta_\mathrm{in}).
\end{eqnarray}
%
This linear combination matches the conventions used in \cite{PhysRevD.83.084002}, and is the frequency-domain equivalent of their Equation~(2.8). The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, and depend on the orientation of detector $i$ as well as the declination and local hour angle $h_i$ of the source, and are derived in~\cite{PhysRevD.63.042003}. The quantity $D_{1,i}$ is a fiducial horizon distance for detector $i$,
%
\begin{equation}\label{eq:horizon}
D_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|H_0(\omega; \mathbf \theta)\right|^2}{S_i(\omega)} \,\mathrm{d}\omega.
\end{equation}

@@

More succinctly, we can write the signal received by detector $i$ in terms of direct observables, the amplitude $\rho_i$, phase $\gamma_i$, and time delay $\tau_i$ on arrival at detector $i$,
%
\begin{equation}\label{eq:signal-model}
	H_i (\omega; \boldsymbol\theta) = \frac{\rho_i}{\sigma_i} e^{i (\gamma_i - \omega \tau_i)} H_0(\omega; \boldsymbol\theta_\mathrm{in}).
\end{equation}

The prevailing technique for detection of \acp{GW} from \acp{CBC} is to realize a \ac{MLE} from \Eref{eq:gaussian-likelihood} with the above signal model. This takes the form of a bank of matched filters, or template waveforms that are continually convolved with the incoming data stream. The matched filter or \ac{ML} estimates are given by
%
\begin{eqnarray}
\hat{\boldsymbol\theta}_i, \{\hat\tau_i\}_i &=& \argmax_{\boldsymbol\theta_\mathrm{in}, \{\hat\tau_i\}_i} \sum_i \frac{1}{{\sigma_i}^2} \left| \int_0^\infty \frac{{H_0}^*(\omega; \boldsymbol\theta_\mathrm{in}) X_i(\omega) e^{-i \omega \tau_i}}{S_i(\omega)} \,\mathrm{d}\omega \right|^2 \\
\hat\rho_i &=& \frac{1}{\sigma_i} \left| \int_0^\infty \frac{{H_0}^*(\omega; \hat{\boldsymbol\theta}_\mathrm{in}) X_i(\omega) e^{-i \omega \hat\tau_i}}{S_i(\omega)} \,\mathrm{d}\omega \right| \\
\hat\gamma_i &=& \arg \int_0^\infty \frac{{H_0}^*(\omega; \hat{\boldsymbol\theta}_\mathrm{in}) X_i(\omega) e^{-i \omega \hat\tau_i}}{S_i(\omega)} \,\mathrm{d}\omega.
\end{eqnarray}

@@

In the Bayesian framework, the parameters are inferred from the data by forming the posterior distribution, $p(\boldsymbol\theta|\mathbf X)$, which describes the probability of the parameters given the observations. Bayes' rule relates the likelihood $p(\mathbf X|\boldsymbol\theta)$ to the posterior $p(\boldsymbol\theta|\mathbf X)$,
%
\begin{equation}\label{bayes}
p(\boldsymbol\theta|\mathbf X) = \frac{p(\mathbf X|\boldsymbol\theta) p(\boldsymbol\theta)}{p(\mathbf X)},
\end{equation}
%
introducing the prior distribution $p(\boldsymbol\theta)$ which encapsulates previous information about the paramters (for example, arising from earlier observations or from known physical bounds on the parameters), and the evidence $p(\mathbf X)$ which can be thought of as a normalization factor or as describing the parsimoniousness of the model.

@@

In Bayesian inference, although it is often easy to write down the likelihood or even the full posterior in closed form, usually one is interested in only a subset $\boldsymbol\beta$ of all of the model's parameters, the others $\boldsymbol\lambda$ being nuisance parameters. In this case, we integrate away the nuisance parameters, forming the marginal posterior
%
\begin{equation}\label{eq:marginal-posterior}
	p(\boldsymbol\beta|\mathbf X) = \int \frac{p(\mathbf X|\boldsymbol\beta,\boldsymbol\lambda) p(\boldsymbol\beta,\boldsymbol\lambda)}{p(\mathbf X)} \,\mathrm{d}\boldsymbol\lambda
\end{equation}
%
with $\boldsymbol\theta = (\boldsymbol\beta, \boldsymbol\lambda)$. For instance, for the purpose of locating a \ac{GW} source on the sky, all parameters but $(\alpha, \delta)$ are nuisance parameters.

@@

In the frequency domain, the strain $X (\omega)$ observed by a detector is a linear combination of two orthogonal \ac{GW} quadratures and additive noise,
%
\begin{equation}\label{eq:signal-model}
	X (\omega) = H (\omega, \rho, \gamma, \tau) + N (\omega) = \frac{\rho}{\sigma} e^{i (\gamma - \omega \tau)} U (\omega) + N (\omega),
\end{equation}
%
where $H (\omega)$ is the signal content, $N (\omega)$ is the detector's \ac{WSS} Gaussian noise with \ac{PSD} $S(\omega)$, $U (\omega)$ is the frequency-domain post-Newtonian inspiral signal, $\rho$ is the signal's amplitude, $\gamma$ is a phase angle the encodes the relative strength of two \ac{GW} quadratures, and $\tau$ is the time of arrival relative to an arbitrary time base. The quantity $\sigma$ is a normalization constant defined as the \ac{SNR} of the \ac{GW} source at a fiducial distance, here taken to be $D_\mathrm{fiducial} = 1$~Mpc,
%
\begin{equation}\label{eq:sigma}
    \sigma^2 = \int_0^\infty \frac{\left| U (\omega)\right|^2}{S(\omega)} \, \mathrm{d}\omega.
\end{equation}
%
The distance at which an optimally oriented \ac{GW} source would produce \ac{SNR}=1 is defined as $D_1 = D_\mathrm{fiducial} / \sigma$. The likelihood, or the probability of obtaining the observation $X(\omega)$ conditioned on the value of the parameter vector $\boldsymbol\theta = (\rho, \gamma, \tau)$, is Gaussian, with the log-likelihood proportional to:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\mathcal{L}(X; \boldsymbol\theta) = p(X | \boldsymbol\theta)
		\propto \exp \left[
		- \frac{1}{2} \int_0^\infty \frac{\left|X (\omega)
			- H(\omega; \boldsymbol\theta) \right|^2}{S(\omega)} \, \mathrm{d}\omega
	\right].
\end{equation}

The Fisher information matrix for a measurement $\tilde x$ described by the unknown parameter vector $\boldsymbol{\theta}$ is
%
\begin{equation}\label{eq:general-fisher-matrix}
	\mathcal{I}_{jk} = \mathrm{E} \, \left[
		-\frac{\partial^2 \log
			\mathcal{L}(X ; \boldsymbol\theta)}
			{\partial \theta_j \theta_k}
	\right].
\end{equation}
%
This equation involves double derivatives, but substantial simplification is possible when---as in this case---the likelihood is Gaussian:
%
\begin{equation}\label{eq:gaussian-fisher-matrix}
	\mathcal{I}_{jk} = \int_0^\infty \Re \left[
        \left(\frac{\partial H}{\partial \theta_j}\right)^*
        \left(\frac{\partial H}{\partial \theta_k}\right)
	\right] \frac{1}{S(\omega)} \, \mathrm{d}\omega.
\end{equation}
%
This form is more useful because it involves only first derivatives, and of the signal $\tilde h (\omega)$ rather than the entire observation $\tilde x (\omega)$. In terms of the $k$th \ac{SNR}-weighted moment of angular frequency,
%
\begin{equation}\label{eq:angular-frequency-moments}
    \overline{\omega^k} =
        \left[ \int_0^\infty \frac{|U (\omega)|^2}{S(\omega)} \omega^k \, \mathrm{d}\omega \right]
        \left[ \int_0^\infty \frac{|U (\omega)|^2}{S(\omega)} \, \mathrm{d}\omega \right]^{-1},
\end{equation}
%
the Fisher matrix for the signal model in \Eref{eq:signal-model} is
%
\begin{equation}\label{eq:fisher-matrix}
	\mathcal{I} = \bordermatrix{
        ~ & \rho & \gamma & \tau \cr
        \rho & 1 & 0 & 0 \cr
        \gamma & 0 & \rho^2 & -\rho^2 \overline{\omega} \cr
        \tau & 0 & -\rho^2 \overline{\omega} & \rho^2 \overline{\omega^2}
    }.
\end{equation}

>>>>

A laser interferometer senses \acp{GW} by transducing differential changes in arm length to a digitized photocurrent time series. In the angular frequency domain, a single detector's observation $X_i(\omega)$ is
%
\begin{equation}\label{eq:signal-model}
	x_i (\omega) = h_i (\omega; \boldsymbol\theta) + n_i (\omega),
\end{equation}
%
where $h_i (\omega; \boldsymbol\theta)$ is the \ac{GW} signal receied by detector $i$, given a parameter vector $\boldsymbol\theta$ that describes the \ac{GW} source, and $n_i (\omega)$ is that detector's \ac{WSS} Gaussian noise with \ac{PSD} $S_i(\omega)$. We shall denote the combined observation from a network of detectors as $\mathbf x (\omega) \equiv \{x_i (\omega)\}_i$.

The likelihood, or the probability of obtaining the observation $\mathbf x$ conditioned on the value of $\boldsymbol\theta$, is Gaussian, with the log\nobreakdashes-likelihood proportional to:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\mathcal{L}(\mathbf x; \boldsymbol\theta) = \prod_i p(x_i | \boldsymbol\theta)
		\propto \exp \left[
		- \frac{1}{2} \sum_i \int_0^\infty \frac{\left|x_i (\omega)
			- h_i(\omega; \boldsymbol\theta) \right|^2}{S_i(\omega)} \, \mathrm{d}\omega
	\right].
\end{equation}

@@

Assuming circular orbits and no spin precession, we can write the \ac{GW} signal in each detector as a linear combination of two basis waveforms $h_0$ and $h_{\pi/2}$. For nonprecessing systems, $h_0$ and $h_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. In other words, $h_{\pi/2}(\omega) = i h_0(\omega)$ for all $\omega$. This linear combination isolates all dependence on the extrinsic parameters $\boldsymbol\theta_\mathrm{ex}$ into the coefficients and all dependence on the intrinsic parameters $\boldsymbol\theta_\mathrm{in}$ into the basis waveform,
%
\begin{eqnarray}\label{eq:full-signal-model}
	\fl h_i(\omega; \boldsymbol\theta) &=& e^{-i\omega t_\oplus}\frac{D_{1,i}}{D_\mathrm{L}}
	\left(\begin{array}{cc}
		F_{+,i}(h_i, \delta) \\
		F_{\times,i}(h_i, \delta)
	\end{array}\right)^\intercal
	\left(\begin{array}{cc}
		\cos{2\psi} & -\sin{2\psi} \\
		\sin{2\psi} & \cos{2\psi}
	\end{array}\right)
	\left(\begin{array}{cc}
		\frac{1}{2}\left(1+\cos^2{\iota}\right) & 0 \\
		0 & \cos{\iota}
	\end{array}\right) \nonumber\\\fl &\cdot&
	\left(\begin{array}{cc}
		\cos{2\phi_c} & -\sin{2\phi_c} \\
		\sin{2\phi_c} & \cos{2\phi_c}
	\end{array}\right)
	\left(\begin{array}{cc}
		1 \\
		i
	\end{array}\right) h_0(\omega; \boldsymbol\theta_\mathrm{in}).
\end{eqnarray}
%
This linear combination matches the conventions used in \cite{PhysRevD.83.084002}, and is the frequency-domain equivalent of their Equation~(2.8). The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, and depend on the orientation of detector $i$ as well as the declination and local hour angle $h_i$ of the source, and are derived in~\cite{PhysRevD.63.042003}. The quantity $D_{1,i}$ is a fiducial horizon distance for detector $i$,
%
\begin{equation}\label{eq:horizon}
D_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|h_0(\omega; \mathbf \theta)\right|^2}{S_i(\omega)} \,\mathrm{d}\omega.
\end{equation}

@@

More succinctly, we can write the signal received by detector $i$ in terms of direct observables, the amplitude $\rho_i$, phase $\gamma_i$, and time delay $\tau_i$ on arrival at detector $i$,
%
\begin{equation}\label{eq:signal-model}
	h_i (\omega; \boldsymbol\theta) = \frac{\rho_i}{\sigma_i} e^{i (\gamma_i - \omega \tau_i)} h_0(\omega; \boldsymbol\theta_\mathrm{in}).
\end{equation}

The prevailing technique for detection of \acp{GW} from \acp{CBC} is to realize a \ac{MLE} from \Eref{eq:gaussian-likelihood} with the above signal model. This takes the form of a bank of matched filters, or template waveforms that are continually convolved with the incoming data stream. The matched filter or \ac{ML} estimates are given by
%
\begin{eqnarray}
\hat{\boldsymbol\theta}_i, \{\hat\tau_i\}_i &=& \argmax_{\boldsymbol\theta_\mathrm{in}, \{\hat\tau_i\}_i} \sum_i \frac{1}{{\sigma_i}^2} \left| \int_0^\infty \frac{{h_0}^*(\omega; \boldsymbol\theta_\mathrm{in}) x_i(\omega) e^{-i \omega \tau_i}}{S_i(\omega)} \,\mathrm{d}\omega \right|^2 \\
\hat\rho_i &=& \frac{1}{\sigma_i} \left| \int_0^\infty \frac{{h_0}^*(\omega; \hat{\boldsymbol\theta}_\mathrm{in}) x_i(\omega) e^{-i \omega \hat\tau_i}}{S_i(\omega)} \,\mathrm{d}\omega \right| \\
\hat\gamma_i &=& \arg \int_0^\infty \frac{{h_0}^*(\omega; \hat{\boldsymbol\theta}_\mathrm{in}) x_i(\omega) e^{-i \omega \hat\tau_i}}{S_i(\omega)} \,\mathrm{d}\omega.
\end{eqnarray}

@@

In the Bayesian framework, the parameters are inferred from the data by forming the posterior distribution, $p(\boldsymbol\theta|\mathbf x)$, which describes the probability of the parameters given the observations. Bayes' rule relates the likelihood $p(\mathbf x|\boldsymbol\theta)$ to the posterior $p(\boldsymbol\theta|\mathbf x)$,
%
\begin{equation}\label{bayes}
p(\boldsymbol\theta|\mathbf x) = \frac{p(\mathbf x|\boldsymbol\theta) p(\boldsymbol\theta)}{p(\mathbf x)},
\end{equation}
%
introducing the prior distribution $p(\boldsymbol\theta)$ which encapsulates previous information about the paramters (for example, arising from earlier observations or from known physical bounds on the parameters), and the evidence $p(\mathbf x)$ which can be thought of as a normalization factor or as describing the parsimoniousness of the model.

@@

In Bayesian inference, although it is often easy to write down the likelihood or even the full posterior in closed form, usually one is interested in only a subset $\boldsymbol\beta$ of all of the model's parameters, the others $\boldsymbol\lambda$ being nuisance parameters. In this case, we integrate away the nuisance parameters, forming the marginal posterior
%
\begin{equation}\label{eq:marginal-posterior}
	p(\boldsymbol\beta|\mathbf x) = \int \frac{p(\mathbf x|\boldsymbol\beta,\boldsymbol\lambda) p(\boldsymbol\beta,\boldsymbol\lambda)}{p(\mathbf x)} \,\mathrm{d}\boldsymbol\lambda
\end{equation}
%
with $\boldsymbol\theta = (\boldsymbol\beta, \boldsymbol\lambda)$. For instance, for the purpose of locating a \ac{GW} source on the sky, all parameters but $(\alpha, \delta)$ are nuisance parameters.

@@

In the frequency domain, the strain $X (\omega)$ observed by a detector is a linear combination of two orthogonal \ac{GW} quadratures and additive noise,
%
\begin{equation}\label{eq:signal-model}
	x (\omega) = h (\omega, \rho, \gamma, \tau) + n (\omega) = \frac{\rho}{\sigma} e^{i (\gamma - \omega \tau)} u (\omega) + n (\omega),
\end{equation}
%
where $h (\omega)$ is the signal content, $n (\omega)$ is the detector's \ac{WSS} Gaussian noise with \ac{PSD} $S(\omega)$, $u (\omega)$ is the frequency-domain post-Newtonian inspiral signal, $\rho$ is the signal's amplitude, $\gamma$ is a phase angle the encodes the relative strength of two \ac{GW} quadratures, and $\tau$ is the time of arrival relative to an arbitrary time base. The quantity $\sigma$ is a normalization constant defined as the \ac{SNR} of the \ac{GW} source at a fiducial distance, here taken to be $D_\mathrm{fiducial} = 1$~Mpc,
%
\begin{equation}\label{eq:sigma}
    \sigma^2 = \int_0^\infty \frac{\left| u (\omega)\right|^2}{S(\omega)} \, \mathrm{d}\omega.
\end{equation}
%
The distance at which an optimally oriented \ac{GW} source would produce \ac{SNR}=1 is defined as $D_1 = D_\mathrm{fiducial} / \sigma$. The likelihood, or the probability of obtaining the observation $x(\omega)$ conditioned on the value of the parameter vector $\boldsymbol\theta = (\rho, \gamma, \tau)$, is Gaussian, with the log-likelihood proportional to:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\mathcal{L}(x; \boldsymbol\theta) = p(x | \boldsymbol\theta)
		\propto \exp \left[
		- \frac{1}{2} \int_0^\infty \frac{\left|x (\omega)
			- h(\omega; \boldsymbol\theta) \right|^2}{S(\omega)} \, \mathrm{d}\omega
	\right].
\end{equation}

The Fisher information matrix for a measurement $\tilde x$ described by the unknown parameter vector $\boldsymbol{\theta}$ is
%
\begin{equation}\label{eq:general-fisher-matrix}
	\mathcal{I}_{jk} = \mathrm{E} \, \left[
		-\frac{\partial^2 \log
			\mathcal{L}(x ; \boldsymbol\theta)}
			{\partial \theta_j \theta_k}
	\right].
\end{equation}
%
This equation involves double derivatives, but substantial simplification is possible when---as in this case---the likelihood is Gaussian:
%
\begin{equation}\label{eq:gaussian-fisher-matrix}
	\mathcal{I}_{jk} = \int_0^\infty \Re \left[
        \left(\frac{\partial h}{\partial \theta_j}\right)^*
        \left(\frac{\partial h}{\partial \theta_k}\right)
	\right] \frac{1}{S(\omega)} \, \mathrm{d}\omega.
\end{equation}
%
This form is more useful because it involves only first derivatives, and of the signal $\tilde h (\omega)$ rather than the entire observation $\tilde x (\omega)$. In terms of the $k$th \ac{SNR}-weighted moment of angular frequency,
%
\begin{equation}\label{eq:angular-frequency-moments}
    \overline{\omega^k} =
        \left[ \int_0^\infty \frac{|u (\omega)|^2}{S(\omega)} \omega^k \, \mathrm{d}\omega \right]
        \left[ \int_0^\infty \frac{|u (\omega)|^2}{S(\omega)} \, \mathrm{d}\omega \right]^{-1},
\end{equation}
%
the Fisher matrix for the signal model in \Eref{eq:signal-model} is
%
\begin{equation}\label{eq:fisher-matrix}
	\mathcal{I} = \bordermatrix{
        ~ & \rho & \gamma & \tau \cr
        \rho & 1 & 0 & 0 \cr
        \gamma & 0 & \rho^2 & -\rho^2 \overline{\omega} \cr
        \tau & 0 & -\rho^2 \overline{\omega} & \rho^2 \overline{\omega^2}
    }.
\end{equation}
